Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-17 cordillera-cycle.tex
Add figure or page number in citations.

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent \citep{Clague.1989, Dyke.2004}, directions of past sliding
velocities \citep{Prest.etal.1968, Kleman.etal.2010}, and locations of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.etal.1980, Clague.1981, Porter.Swanson.1998}, cosmogenic dating
\citep[e.g.,][]{Ward.etal.2007, Stroeven.etal.2010, Stroeven.etal.2013}, and
the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle \citep{Clague.1989, Dyke.2004}. However, former ice thickness
and the ice sheet's contribution to the last glacial maximum sea-level
low-stand remain poorly constrained. Moreover, our understanding of the
glaciation history is mainly restricted to the deglaciation phase and a
regional scale, while little is known about the ice sheet evolution prior to
the last glacial maximum extent \citep{Kleman.etal.2010}. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Eurasian and Laurentide and ice sheets
\citep{Kleman.etal.1997, Kleman.etal.2010}, such is not the case for the
Cordilleran ice sheet, where complex arrangements of ice flow directions
emanating from multiple glaciation centres have been identified
\citep{Prest.etal.1968, Clague.1989}, but remain poorly understood
\citep{Kleman.etal.2010}.

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle. Although numerical modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago \citep{Jackson.Clague.1991,Robert.1991}, the ubiquitously mountainous
topography of the region has presented a major challenge to its application.
In fact, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep[e.g.,][]{Bueler.Brown.2009} allows
high-resolution numerical modelling of palaeo-glacier flow on mountainous
terrain over millenial time scales \citep{Golledge.etal.2012}. In the North
American Cordillera, the topographic complexity induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2013}.

Because past climate conditions are subject to considerable uncertainty, our
palaeoclimate forcing is a simplistic approximation of time-evolved temperature
and precipitation fields derived from a combination of a present-day
atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two \chem{\delta^{18}O} records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two \chem{\delta^{18}O}
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores \citep{Herbert.2001}. Model output is compared to
geomorphological evidence in terms of timing and extent of glaciation and
patterns of deglaciation.

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{Bueler.Brown.2009,Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.Blatter.2009,Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{Aschwanden.etal.2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments.
The till effective pressure is related to the modelled amount of water in the
till, using a formula derived from laboratory experiments with till extracted
from an Antarctic ice stream \citep{Tulaczyk.etal.2000}. Basal topography is
derived from the ETOPO1 combined topography and bathymetry dataset with
a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, monthly precipitation, and monthly standard
deviation of near-surface temperature by a~temperature-index (positive
degree-day) model \citep{Hock.2003}. Ice accumulation is equal to precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep{Calov.Greve.2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability
associated with the seasonal cycle itself \citep{Seguinot.Rogozhina.2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study \citep{Seguinot.etal.2013}.

@@

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Lake Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
shore \citep{Herbert.2001}. Palaeo-temperatures from the GRIP and NGRIP records
were calculated using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

However, the timing of the modelled glacial maxima strongly depends on the
choice of palaeo-temperature record used, and a more precise agreement with
timing reconstructed from geological evidence is only attained in some of our
simulations. In the Puget lowland, the last glacial maximum advance of the
southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
\citep{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}.

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning the
simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
deglaciation is caused by an early warming sealed in the palaeo-temperature records
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.2001}. The California current, driving cold waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \citep{Herbert.2001}.

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~kyr after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Original reference\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C
        & \citet{Dansgaard.etal.1993} \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C
        & \citet{Andersen.etal.2004} \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
        & \citet{Jouzel.etal.2007} \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
        & \citet{Petit.etal.1999} \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
        & \citet{Herbert.2001} \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
        & \citet{Herbert.2001} \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

>>>>

For more than a century, exploration and geological investigation of the
Cordillera have led to a large collection of evidence of the former ice cover
\citep{Jackson.Clague.1991}. This evidence consists of mapped boundaries of the
former ice extent (\citealp[Fig.~1.12]{Clague.1989}; \citealp{Duk-Rodkin.1999};
\citealp{Dyke.2004}), directions of past sliding velocities
(\citealp{Prest.etal.1968}; \citealp[Fig.~2]{Kleman.etal.2010}), and locations
of former
melt-water streams \citep{Margold.etal.2011, Margold.etal.2013}, as well as the
timing of glaciation identified from radiocarbon dating \citep[e.g.,][]
{Clague.1981, Porter.Swanson.1998}, cosmogenic dating \citep[e.g.,][]
{Ward.etal.2007, Menounos.etal.2009, Stroeven.etal.2010, Stroeven.etal.2013},
and the offshore sedimental record \citep{Cosma.etal.2008, Davies.etal.2011}.

Field-based evidence allowed to reconstruct, with a high degree of confidence,
the maximal extent attained by the Cordilleran ice sheet during the last
glacial cycle (\citealp[Fig. 1.12]{Clague.1989}; \citealp{Dyke.2004}). However,
former ice thickness
and the ice sheet's contribution to the last glacial maximum sea-level
low-stand remain poorly constrained. Moreover, our understanding of the
glaciation history is mainly restricted to the deglaciation phase and a
regional scale, while little is known about the ice sheet evolution prior to
the last glacial maximum extent \citep[Fig.~6]{Kleman.etal.2010}. Although
time-evolving, whole ice-sheet reconstructions of glacial advance and retreat
patterns are available for the Eurasian and Laurentide and ice sheets
\citep{Kleman.etal.1997, Kleman.etal.2010}, such is not the case for the
Cordilleran ice sheet, where complex arrangements of ice flow directions
emanating from multiple glaciation centres have been identified
\citep[Fig.~1.12]{Prest.etal.1968, Clague.1989}, but remain poorly understood
\citep[p.~2049]{Kleman.etal.2010}.

The present study uses a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle. Although numerical modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet more than twenty years
ago (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991}), the
ubiquitously mountainous
topography of the region has presented a major challenge to its application.
In fact, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep[e.g.,][]{Bueler.Brown.2009} allows
high-resolution numerical modelling of palaeo-glacier flow on mountainous
terrain over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. In the
North
American Cordillera, the topographic complexity induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as input to the ice sheet model
\citep{Seguinot.etal.2013}.

Because past climate conditions are subject to considerable uncertainty, our
palaeoclimate forcing is a simplistic approximation of time-evolved temperature
and precipitation fields derived from a combination of a present-day
atmospheric reanalysis \citep{Mesinger.etal.2006}, lapse-rate corrections,
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two \chem{\delta^{18}O} records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two \chem{\delta^{18}O}
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific oceanic
sediment cores \citep{Herbert.etal.2001}. Model output is compared to
geomorphological evidence in terms of timing and extent of glaciation and
patterns of deglaciation.

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep[Eqn.~15]{Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{PISM-authors.2014} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments.
The till effective pressure is related to the modelled amount of water in the
till, using a formula derived from laboratory experiments with till extracted
from an Antarctic ice stream \citep[Eqn.~2]{Tulaczyk.etal.2000}. Basal
topography is
derived from the ETOPO1 combined topography and bathymetry dataset with
a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

@@

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, monthly precipitation, and monthly standard
deviation of near-surface temperature by a~temperature-index (positive
degree-day) model \citep[e.g.,]{Hock.2003}. Ice accumulation is equal to
precipitation
when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly
with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed
from the number of positive degree-days, defined as the integral of
temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep[Eqn.~6]{Calov.Greve.2005} is
numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, after excluding variability
associated with the seasonal cycle itself \citep[cf.][]
{Seguinot.Rogozhina.2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006} over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study \citep{Seguinot.etal.2013}.

@@

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Lake Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
shore \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and NGRIP
records were calculated using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

However, the timing of the modelled glacial maxima strongly depends on the
choice of palaeo-temperature record used, and a more precise agreement with
timing reconstructed from geological evidence is only attained in some of our
simulations. In the Puget lowland, the last glacial maximum advance of the
southern Cordilleran ice sheet margin has been constrained by radiocarbon
dating between 17.4 and 16.4\,\unit{\chem{^{14}C}\,cal\,kyr\,BP}
\citep[Fig.~4]{Porter.Swanson.1998}.
These dates have been confirmed by the offshore sediment record, which shows an
increase of glaciomarine sedimentation between 19.5 and
16.2\,\unit{\chem{^{14}C}\,cal\,kyr\,BP} \citep{Cosma.etal.2008}.

Among simulations presented here, only those forced by the GRIP, EPICA and
Vostok records place the last glacial maximum within this range
(Fig.~\ref{fig:timeseries}, lower panel). Simulations driven by the NGRIP,
ODP~1012 and ODP~1020 records, on the contrary, produce a last glacial maximum
that pre-dates reconstructed ages by several thousand years. Concerning the
simulations driven by oceanic (ODP~1012, ODP~1020) records, this early
deglaciation is caused by an early warming sealed in the palaeo-temperature records
(Fig.~\ref{fig:timeseries}, upper panel). However, it has been shown that this
early warming is a local effect, corresponding to a weakening of the California
current \citep[Fig.~3]{Herbert.etal.2001}. The California current, driving cold
waters southwards
along the western coast of North America, is thought to have weakened in the
presence of the Cordilleran ice sheet, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
during the last glacial maximum \citep{Herbert.etal.2001}.

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~kyr after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Original reference\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C
        & \citet{Dansgaard.etal.1993} \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C
        & \citet{Andersen.etal.2004} \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
        & \citet{Jouzel.etal.2007} \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
        & \citet{Petit.etal.1999} \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
        & \citet{Herbert.etal.2001} \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
        & \citet{Herbert.etal.2001} \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}
